Biostat III exercises in R
===========
Laboratory exercise 1b
-----------

### Suggested solutions by
Author: Annika Tillander, 2014-01-30 <br/>
Edited: Andreas Karlsson, 2015-02-27, 2016-03-01; Mark Clements, 2017-08-02


-----------

```{r setup, cache=FALSE, message=FALSE, echo=FALSE}
library('knitr')
read_chunk('q1.R')
opts_chunk$set(cache=FALSE, fig.width=7, fig.height=6)
```

You may have to install the required packages the first time you use
them. You can install a package by
`install.packages("package_of_interest")` for each package you
require.

```{r loadDependecies, message=FALSE}
```

Life tables are available using the `lifetab` function from the `KMsurv` package on CRAN. We have written a small wrapper `lifetab2` which allows for a `Surv` object and a dataset.
Following are the life table estimates. Note that in the lectures,
when we estimated all-cause survival, there were 8 deaths in the first
interval. One of these died of a cause other than cancer so in the
cause-specific survival analysis we see that there are 7 ‘deaths’ and
1 censoring (in `lifetab2` we used the term ‘nlost’ for number lost to
follow-up) in the first interval.

```{r lifeTable, message=FALSE}
```

Following is a table of Kaplan-Meier estimates. It is not clear from
the table, if the person censored at time 2 was at risk when the other
person dies at time 2. Following the table is a graph of the survival
function.


```{r KaplanMeier, message=FALSE}
```
