Biostat III exercises in R
===========
Laboratory exercise 12
-----------

### Suggested solutions by
Author: Andreas Karlsson, 2015-03-06 <br/>


## Cox model for cause-specific mortality for melanoma (all stages)


Use Cox regression to model the cause-specific survival of patients with skin melanoma (including all stages).

-----------


```{r setup, cache=FALSE, message=FALSE, echo=FALSE}
library('knitr')
read_chunk('q12.R')
opts_chunk$set(cache=FALSE)
```

You may have to install the required packages the first time you use them. You can install a package by `install.packages("package_of_interest")` for each package you require.

```{r loadDependecies, message=FALSE}
```
Load the melanoma data and explore it.

```{r loadPreprocess, results='hide'}
```

**(a)** First fit the model with sex as the only explanatory variable. Does there appear to be a difference in survival between males and females?


```{r 12.a, warning=FALSE, message=FALSE}
```

**(b)** Is the effect of sex confounded by other factors (e.g. age, stage, subsite, period)? After controlling for potential confounders, does there still appear to a difference in survival between males and females?

```{r 12.b, warning=FALSE, message=FALSE}
```

**(c)** Consider the hypothesis that there exists a class of melanomas where female sex hormones play a large role in the etiology. These hormone related cancers are diagnosed primarily in women and are, on average, less aggressive (i.e., prognosis is good). If such a hypothesis were true we might expect the effect of sex to be modified by age at diagnosis (e.g., pre versus post menopausal). Test whether this is the case.

```{r 12.c, warning=FALSE, message=FALSE}
```

**(d)** Decide on a ‘most appropriate’ model for these data. Be sure to evaluate the proportional hazards assumption.


```{r 12.d, warning=FALSE, message=FALSE}
```
