---
title: "Biostatistics III in R"
author:
- Code by Annika Tillander, Andreas Karlsson and Mark Clements
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

# Exercise 1: Life tables and Kaplan-Meier estimates of survival #

```{r setup, cache=FALSE, message=FALSE, echo=FALSE}
library('knitr')
read_chunk('../q1.R')
opts_chunk$set(cache=FALSE, fig.width=7, fig.height=6)
options(width=90)
```

## (a) Hand calculation: Life table and Kaplan-Meier estimates of survival ##

The results are contained in an [Excel file](exercise1.xls) and are also shown in the R
    output below.

## (b) Using R to validate the hand calculations done in part 1 (a) ##

First, load the `biostat3` library:

```{r loadDependencies, message=FALSE}
```

Following are the life table estimates. Note that in the lectures,
    when we estimated all-cause survival, there were 8 deaths in the
    first interval. One of these died of a cause other than cancer so in
    the cause-specific survival analysis we see that there are 7
    ‘deaths’ and 1 censoring (Stata uses the term ‘lost’ for lost
    to follow-up) in the first interval.

```{r lifeTable}
```

Following is a table of Kaplan-Meier estimates. Although it’s not
    clear from the table, the person censored (lost) at time 2 was at
    risk when the other person dies at time 2. On the following is
    a graph of the survival function.

```{r KaplanMeier}
```

