---
title: "Biostatistics III in R"
author:
- Code by Johan Zetterqvist and Mark Clements
output:
  html_document:
    theme: null
  html_vignette: default
  prettydoc::html_pretty: default
---

# Exercise 11. Cox regression with all-cause mortality as the outcome #


Now fit a model to the localised melanoma data where the outcome is observed survival (i.e. all deaths are considered to be events).

```{r setup, cache=FALSE, message=FALSE, echo=FALSE}
library('knitr')
read_chunk('../q11.R')
opts_chunk$set(cache=FALSE)
```

You may have to install the required packages the first time you use them. You can install a package by `install.packages("package_of_interest")` for each package you require.

```{r loadDependencies, message=FALSE}
```
Load the melanoma data and explore it.

```{r loadPreprocess, results='hide'}
```

## (a) ## 

For patients of the same sex diagnosed in the same period, those
        aged 60–74 at diagnosis have a 2.9 times higher risk of death
        *due to any causes* than those aged 0–44
        at diagnosis. This difference is statistically significant.

```{r 11.a, warning=FALSE}
```

## (b) ## 

Note that the previous model estimated cause-specific hazard
        ratios whereas the current model estimates all-cause
        hazard ratios. The estimated hazard ratios for sex and period
        are similar, whereas the estimated hazard ratios for age are
        markedly different. This is because non-cancer mortality is
        heavily dependent on age, but only lightly dependent on sex and
        calendar period.
		
```{r 11.b, warning=FALSE}
```
