---
title: "Biostatistics III in R"
author:
- Code by Johan Zetterqvist and Mark Clements
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

# Exercise 11. Cox regression with all-cause mortality as the outcome #


Now fit a model to the localised melanoma data where the outcome is observed survival (i.e. all deaths are considered to be events).

```{r setup, cache=FALSE, message=FALSE, echo=FALSE}
library('knitr')
read_chunk('../q11.R')
opts_chunk$set(cache=FALSE)
```

You may have to install the required packages the first time you use them. You can install a package by `install.packages("package_of_interest")` for each package you require.

```{r loadDependencies, message=FALSE}
```
Load the melanoma data and explore it.

```{r loadPreprocess, results='hide'}
```

## (a) ## 

For patients of the same sex diagnosed in the same period, those
aged 60–74 at diagnosis have a 2.94 times higher risk of death
*due to any causes* than those aged 0-–44
at diagnosis. The 95% confidence interval is (2.52, 3.44). This difference is statistically significant.

The regression equation is 
\begin{align*}
h(t|\text{year8594},\text{sex},\text{agegrp}) &= h_0(t) \exp(\beta_1 I(\text{sex}=\text{"Female"})+\beta_2 I(\text{year8594}=\text{"Diagnosed 85-94"})+\\
&\qquad\beta_3 I(\text{agegrp}=\text{"45-59"})+\beta_4 I(\text{agegrp}=\text{"60-64"}) + \beta_5 I(\text{agegrp}=\text{"75+"}) 
\end{align*}
where $h(t|\text{year8594},\text{sex},\text{agegrp},\text{agegrp})$ is the hazard at time $t$ given covariates $\text{year8594}$, $\text{sex}$ and $\text{agegrp}$ and $\text{agegrp}$, with baseline hazard $h_0(t)$ and regression coefficients representing log hazard ratios for $\beta_1$ for the calendar period 1985--1994, $\beta_2$ for females, with log hazard ratios at time 0 for those aged 45--59 years, 60--74 years and 75 years and over are $\beta_3$, $\beta_4$ and $\beta_5$, respectively.

From the regression equation, a formula for the hazard ratio comparing a male aged 60--74 years with a female aged 75 years and over is
\begin{align*}
\frac{h(t|\text{year8594},\text{sex}=\text{"Male"},\text{agegrp}=\text{"60-74"})}{h(t|\text{year8594},\text{sex}=\text{"Female"},\text{agegrp}=\text{"75+"})} &= 
\frac{h_0(t) \exp(\beta_2 I(\text{year8594}=\text{"Diagnosed 85-94"})+ \beta_4)}{h_0(t)\exp(\beta_1}+\beta_2 I(\text{year8594}=\text{"Diagnosed 85-94"}+\beta_5))} \\
&= \exp(\beta_4 - \beta_1 - \beta_5)
\end{align*}

```{r 11.a, warning=FALSE}
```

## (b) ## 

Note that the previous model estimated cause-specific hazard
ratios whereas the current model estimates all-cause
hazard ratios. The estimated hazard ratios for sex and period
are similar, whereas the estimated hazard ratios for age are
markedly different. This is because non-cancer mortality is
heavily dependent on age, but only lightly dependent on sex and
calendar period.
		
```{r 11.b, warning=FALSE}
```
