---
title: "Biostatistics III in R"
author:
- Code by Annika Tillander, Andreas Karlsson and Mark Clements
output:
  html_document:
    theme: null
  html_vignette: default
  prettydoc::html_pretty: default
---

# Exercise 3. Localised melanoma: Comparing estimates of cause-specific survival between periods; first graphically and then using the log rank test #


```{r setup, cache=FALSE, message=FALSE, echo=FALSE}
library('knitr')
read_chunk('../q3.R')
opts_chunk$set(cache=FALSE, fig.width=10, fig.height=6)
```

You may have to install the required packages the first time you use them. You can install a package by `install.packages("package_of_interest")` for each package you require.

```{r loadDependencies, message=FALSE}
```

We will now analyse the full data set of patients diagnosed with localised skin melanoma. We start by reading the data selecting those with a localised stage and then define a 1/0 varible for the events that we are interested in.

```{r loadPreprocess}
```

## (a) ##

```{r a_survDiaDate}
```

There seems to be a clear difference in survival between the
        two periods. Patients diagnosed during 1985–94 have superior
        survival to those diagnosed 1975–84.

## (b) ##

The plot shows the instantaneous cancer-specific mortality rate
        (the hazard) as a function of time. It appears that mortality is
        highest approximately 40 months following diagnosis. Remember
        that all patients were classified as having localised cancer at
        the time of diagnosis so we would not expect mortality to be
        high directly following diagnosis.

The plot of the hazard clearly illustrates the pattern of
        cancer-specific mortality as a function of time whereas this
        pattern is not obvious in the plot of the survivor function.

```{r b_hazDiaDate}
```


## (c) ##

There is strong evidence that survival differs between the
        two periods. The log-rank and the Wilcoxon tests give very
        similar results. The Wilcoxon test gives more weight to
        differences in survival in the early period of follow-up (where
        there are more individuals at risk) whereas the log rank test
        gives equal weight to all points in the follow-up. Both tests
        assume that, if there is a difference, a proportional hazards
        assumption is appropriate.

```{r c_testDiaDate}
```


## (d) ##
We see that mortality increases with age at diagnosis (and
        survival decreases).
		
```{r d_crudeRates1000_agegrp}
```

The rates are (cause-specific) deaths per 1000 person-months.
        When we stset we defined time as time in months and then asked
        for rates per 1000 units of time.

```{r d_plotSurv_sex}
```

What are the units of the estimated hazard rates? HINT: look at how you defined time.

## (e) ##

```{r e_crudeRates1000_agegrp}
```

## (f) ##

```{r f_sexDiff}
```
