---
title: "Biostatistics III in R"
author:
- Code by Annika Tillander, Andreas Karlsson, Benedicte Delcoigne, Johan Zetterqvist and Mark Clements
---


# Exercise 8. Diet data: Using Poisson regression to study the effect of energy intake adjusting for confounders on two different timescales #

Use Poisson regression to study the association between energy intake (hieng) and CHD adjusted for potential confounders (job, BMI). We know that people who expend a lot of energy (i.e., are physically active) require a higher energy intake. We do not have data on physical activity but we are hoping that occupation (job) will serve as a surrogate measure of work-time physical activity (conductors on London double-decker busses expend energy walking up and down the stairs all day).

Fit models both without adjusting for ‘time’ and by adjusting for attained age (you will need to split the data) and time-since-entry and compare the results.

-----------

Load the diet data using time-on-study as the timescale.

```{r setup, cache=FALSE, message=FALSE, echo=FALSE}
library('knitr')
read_chunk('../q8.R')
opts_chunk$set(cache=FALSE)
```

You may have to install the required packages the first time you use them. You can install a package by `install.packages("package_of_interest")` for each package you require.

```{r loadDependencies, message=FALSE}
```
Load diet data and explore it.

```{r loadPreprocess, results='hide'}
```

The rates plotted on timescale attained age show a clear
        increasing trend as age increases, which is to be expected
        (older persons are more likely to suffer from CHD). The rates
        plotted on timescale time-since-entry are almost constant (if
        you have some imagination you can see that the rates are flat).
		

## (a) ##

```{r 8a1_Haz_att_age,  message=FALSE, warning = FALSE}
```

```{r 8a2_Haz_time_entry,  message=FALSE, warning = FALSE}
```

## (b) ##  

Patients with high energy intake have 48% less CHD rate. The
        underlying shape of the rates is assumed to be constant (i.e.
        the baseline is flat) over time.
		
```{r 8b_ir,  message=FALSE}
```

## (c) ##  

The effect of high energy intake is slightly confounded by bmi
        and job, since the point estimate changes a little.


```{r 8c_ir,  message=FALSE}
```

## (d) ##  

The y variable is not correct since it is kept for all splitted
        records, and contains the complete follow-up rather than the
        risktime in that specific timeband. The risktime variable
        contains the correct amount of risktime for each timeband. The
        event variable chd is not correct since it is kept constant
        for all splitted records, while it should only be 1 for the
        last record (if the person has the event). For all other
        records (timebands) for that person it should be 0.
		

```{r 8d_agebands,  message=FALSE}
```

The effect of high energy intake is somewhat confounded by age,
        but also confounded by job and bmi.

```{r 8d_model1,  message=FALSE}
```

Fitting the model for CHD, with adjustment for job and bmi.

```{r 8d_model2,  message=FALSE}
```


## (e) ##
		
```{r 8.e.i,  message=FALSE}
```

There seems to be no confounding by time-since-entry, but there
        is confounding by bmi and job.
