---
title: "Biostatistics III in R"
author:
- Code by Annika Tillander, Johan Zetterqvist and Mark Clements
---

# Exercise 9. Localised melanoma: modelling cause-specific mortality using Cox regression #

In exercise 7 we modelled the cause-specific mortality of patients diagnosed with localised melanoma using Poisson regression. We will now model cause-specific mortality using Cox regression and compare the results to those we obtained using the Poisson regression model.

To fit a Cox proportional hazards model (for cause-specific survival) with calendar period as the only explanatory variable, the following commands can be used. Note that we are censoring all survival times at 120 months (10 years) in order to facilitate comparisons with the Poisson regression model in exercise 7.

-----------


```{r setup, cache=FALSE, message=FALSE, echo=FALSE}
library('knitr')
read_chunk('../q9.R')
opts_chunk$set(cache=FALSE)
```

You may have to install the required packages the first time you use them. You can install a package by `install.packages("package_of_interest")` for each package you require.

```{r loadDependencies, message=FALSE}
```

```{r loadPreprocess, results='hide'}
```

## (a) ##  

```{r 9.a}
```

Patients diagnosed during 1985–94 experience only 77.7% of the
        cancer mortality experienced by those diagnosed 1975–84. That
        is, mortality due to skin melanoma has decreased by 22.3% in the
        latter period compared to the earlier period. This estimate is
        not adjusted for potential confounders. There is strong evidence
        of a statistically significant difference in survival between
        the two periods (based on the test statistic or the fact that
        the CI for the hazard ratio does not contain 1).


## (b) ##  

```{r 9.b}
```

The three test statistics are

        log-rank

        :   14.85 

        Wald

        :   $-3.84^2=14.75$ (from the $z$ test above)

        Likelihood ratio

        :   14.78 (from the output above)

The three test statistics are very similar. We would expect each
        of these test statistics to be similar since they each test the
        same null hypothesis that survival is independent of
        calendar period. The null hypothesis in each case is that
        survival depends on calendar period in such a way that the
        hazard ratio between the two periods is constant over follow-up
        time (i.e. proportional hazards).


## (c) ##  

```{r  9.c}
```

  i. For patients of the same sex diagnosed in the same calendar
            period, those aged 60–74 at diagnosis have an estimated 86%
            higher risk of death due to skin melanoma than those aged
            0–44 at diagnosis. The difference is
            statistically significant.
			
	 If this were an exam question the previous paragraph would
            be awarded full marks. It is worth noting, however, that the
            analysis is adjusted for the fact that mortality may depend
            on time since diagnosis (since this is the underlying
            time scale) and the mortality ratio between the two age
            groups is assumed to be the same at each point during the
            follow-up (i.e., proportional hazard).

  ii. The parameter estimate for period changes from 0.78 to 0.72 when
            age and sex are added to the model. Whether this is
            ‘strong confounding’, or even ‘confounding’, is a matter
            of judgement. I would consider this confounding but not
            strong confounding but there is no correct answer to this
            question.
 
  iii. Age (modelled as a categorical variable with 4 levels) is
            highly significant in the model.


## (d) ##

```{r 9.d, warning=FALSE}
```

Age (modelled as a categorical variable with 4 levels) is highly
        significant in the model. The Wald test is an approximation to
        the LR test and we would expect the two to be similar (which
        they are).

## (e) ##

```{r 9.e}
```

  i. Both models adjust for the same factors. When fitting the
            Poisson regression model we split time since diagnosis into
            annual intervals and explicitly estimated the effect of this
            factor in the model. The Cox model does not estimate the
            effect of ‘time’ but the other estimates are adjusted for
            ‘time’.
			
  ii. Since the two models are conceptually similar we would
            expect the parameter estimates to be similar, which
            they are.
			
  iii. Yes, both models assume ‘proportional hazards’. The
            proportional hazards assumption implies that the risk
            ratios for sex, period, and age are constant across all
            levels of follow-up time. In other words, the assumption
            is that there is no effect modification by follow-up
            time. This assumption is implicit in Poisson regression
            (as it is in logistic regression) where it is assumed that
            estimated risk ratios are constant across all combination
            of the other covariates. We can, of course, relax this
            assumption by fitting interaction terms.

## (f) ##  

```{r 9.f}
```

## (g) ##  

```{r 9.g}
```

