---
title: "Biostatistics III in R"
author:
- Code by Annika Tillander, Andreas Karlsson and Mark Clements
output:
  html_document:
    theme: null
  html_vignette: default
  prettydoc::html_pretty: default
---

# Exercise 1: Life tables and Kaplan-Meier estimates of survival #

## (a) Hand calculation: Life table and Kaplan-Meier estimates of survival ##

Using hand calculation (i.e., using a spreadsheet program or pen,
    paper, and a calculator) estimate the cause-specific survivor
    function for the sample of 35 patients diagnosed with colon
    carcinoma (see the table below) using both the Kaplan-Meier method
    (up to at least 30 months) and the actuarial method (at least the
    first 5 annual intervals).

In the lectures we estimated the observed survivor function
    (i.e. all deaths were considered to be events) using the
    Kaplan-Meier and actuarial methods; your task is to estimate the
    cause-specific survivor function (only deaths due to colon carcinoma
    are considered events) using the same data. The next page includes
    some hints to help you get started.

```{r setup, cache=FALSE, message=FALSE, echo=FALSE}
library('knitr')
read_chunk('../q1.R')
opts_chunk$set(cache=FALSE, fig.width=7, fig.height=6)
options(width=90)
```

```{r loadDependencies, message=FALSE}
```

```{r printData, message=FALSE}
```

### Actuarial approach ###

We suggest you start with the actuarial approach. Your task is to
	construct a life table with the following structure.


    -----  ---  ---  ---  ----  ---  ------
    time   $l$  $d$  $w$  $l'$  $p$  $S(t)$
    -----  ---  ---  ---  ----  ---  ------
    [0-1)   35
	
    [1-2)
	
    [2-3)
	
    [3-4)
	
    [4-5)
	
    [5-6)
    -----  ---  ---  ---  ----  ---  ------

We have already entered $l_1$ (number of people alive at the start
    of interval 1). The next step is to add the number who experienced
    the event ($d$) and the number censored ($w$) during the first year.
    From $l$, $d$, and $w$ you will then be able to calculate $l'$
    (effective number at risk), followed by $p$ (conditional probability
    of surviving the interval) and finally $S(t)$, the cumulative
    probability of surviving from time zero until the end of
    the interval.

    
### Kaplan-Meier approach ###

To estimate survival using the Kaplan-Meier approach you will find
    it easiest to add a line to the table at each and every time there
    is an event or censoring. We should use time in months. The first
    time at which there is an event or censoring is time equal to
    2 months. The trick is what to do when there are both events and
    censorings at the same time.

    ----   ---------  ---  ---  ---  ------
    time   # at risk  $d$  $w$  $p$  $S(t)$
    ----   ---------  ---  ---  ---  ------
	2         35
	
    ----   ---------  ---  ---  ---  ------


    ----   ---------  ---  ---  ---  ------


    ----   ---------  ---  ---  ---  ------



## (b) Using R to validate the hand calculations done in part 1 (a) ##

We will now use R to reproduce the same analyses
    done by hand calculation in section 1.1 although you can do this part
    without having done the hand calculations, since this question also
    serves as an introduction to survival analysis using R. Our aim
    is to estimate the cause-specific survivor function for the sample
    of 35 patients diagnosed with colon carcinoma using both the
    Kaplan-Meier method and the actuarial method. In the lectures we
    estimated the all-cause survivor function (i.e. all deaths were
    considered to be events) using the Kaplan-Meier and actuarial
    methods whereas we will now estimate the cause-specific survivor
    function (only deaths due to colon carcinoma are considered events).


Life tables are available using the `lifetab` function from the `KMsurv` package on CRAN. We have written a small wrapper `lifetab2` which allows for a `Surv` object and a dataset. The following command will give the actuarial estimates:
	
```{r lifeTable, message=FALSE, results="hide"}
```

A listing of the Kaplan-Meier estimates and a graph are obtained as follows

```{r KaplanMeier, message=FALSE, results="hide", fig.show="hide"}
```
